Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-07-04 article/article.tex
Commit 2024-07-05 00:14

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with lower
than average load of aerosols and humidity, which are the main factors
that attenuate the GHI, excluding clouds. With this approach the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. To compensate for this, we
defined the following threshold \(E\) to compare the measured
\(\text{GHI}\): \begin{equation}
\text{CE} : E > 15 + 1.04 \cdot \text{GHI}_\text{ref} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation} This is the criterion of our CE identification.

These values were determined through the implementation of an empirical
method with manual inspection of the CE identification results on
selected days of the whole dataset. We used seven sets of days with
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of EC events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of EC events, (h) randomly selected days, and (i) manually selected
days. For the latter case and where needed, we also used images from a
sky-camera to further aid the decision of the manual inspection.

The definition of the CE events with this method has a degree of
subjectivity, since the actual clear sky irradiance is not known and can
only be approximated. However, this method is capable in detecting all
major CE events. Where some CE events with very low OIR may be not
detected, these are few with small over-irradiance, and it is unlikely
that will affect significantly our results.

@@

\caption{Example of CE identification for 2019-07-11. The green line with blue symbols depicts the measured GHI in one minute steps. The red line shows the modelled threshold $E$ for the detection of CE events, which are denoted with red circles. The black curve represents the TSI at the top of the atmosphere, adjusted for the actual Sun-Earth distance and multiplied by the cosine of the cosine of SZA.}\label{fig:example-day}
\end{figure}

@@

\caption{Example scatter plot of the measured GHI and the reference clear sky irradiance for the year 2005. The over-irradiance for CE and ECE events is color coded, while the remaining data points are shown in black.}\label{fig:example-year}
\end{figure}

@@

The main focus of this study is to investigate the time evolution of the
CE events by analyzing the GHI measurements at Thessaloniki. Cloud
enhancements can be influenced by different factors, such as the
geometry, size and optical thickness of clouds, their height in the
atmosphere and on local weather regimes
\citep{Mol2023, Veerman2022, Gristey2022, Tzoumanikas2016}. Some of
these factors are related to changes in climate; hence it would be
reasonable to expect capturing their contributions to the frequency of
occurrence of CE events over Thessaloniki, as well as to the average OIR
and excess irradiation. The long-term trends were calculated using a
first-order autoregressive model with the `maximum likelihood' fitting
method \citep{Gardner1980, Jones1980}, by implementing the function
`arima' from the library `stats' of the R programming language
\citep{RCT2023}. All trends are reported together with their \(2\sigma\)
error.

@@

Another interesting aspect of the CE events is their distribution within
the year. Figure\nobreakspace{}\ref{fig:relative-month-occurrences}
shows the monthly box and whisker plot of the CE number of occurrence
normalized with the highest median value, that occurs in June, depicting
a clear seasonal cycle. Although CE events are present throughout the
year, the most active months are May and June. During the winter
(December -- February), the number of CE cases is about \(25\,\%\) of
the maximum, while in the intermediate months, the number of occurrences
gradually ramps between the maximum and minimum. This seasonality is a
combined effect of different factors, among them the types of clouds,
their frequency of occurrence, the seasonally varying relative position
of the sun, as well as the local landscape characteristics that may
influence the formation of the clouds. Unfortunately, lack of detailed
data on cloud formation, type and location is not allowing further
analysis. The interannual variability of the monthly CE events is quite
high as manifested by the large monthly extremes, especially in the
summer.

@@

\caption{Seasonal variability of the number of CE events normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have been also normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles, the thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values and the solid circles are outliers.}\label{fig:relative-month-occurrences}
\end{figure}

@@

\caption{Relative frequency distribution of CE events as a function of OIR.}\label{fig:ovir-distribution}
\end{figure}

@@

The duration of the CE events is variable and can last for several
minutes or even more than an hour. In order to study the characteristics
of these consecutive events, we grouped them into bins of increasing
duration in steps of one minute. We have identified 28468 groups of CE
in the whole period of study, where the group of the longest duration of
140 minutes occurred on 07~July 2013.
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the
frequency distribution of the CE events according to their duration. We
conclude that although some groups of events last for more than an hour
about \(80\,\%\) have duration of less than 5 minutes.

@@

\caption{Relative frequency distribution of CE groups of consequent CE events according to their duration.}\label{fig:ceg-duration-distribution}
\end{figure}

The relation between the duration and the mean OIR of the groups has
also been studied (Figure\nobreakspace{}\ref{fig:group-2d}). Evidently,
evens of high excess irradiation have small duration and vice versa. The
vast majority of grouped events are associated with small excess
irradiation (e.g.~\(<5\,\text{kJ}/\text{m}^2\)) and small duration
(e.g.~\(<5\,\text{min}\)) while groups with strong excess irradiation
and long duration are very rare. Similar results of this relation have
been reported by \citet{Zhang2018}, on a study using a far higher
sampling rate than ours.

@@

\caption{Relation of excess irradiation of CE groups with their duration. The logarithmic color scale denotes the frequency of the respective groups of events.}\label{fig:group-2d}
\end{figure}

@@

An aspect of the CE events that is commonly reported and has some
significance on the solar energy production infrastructure are the
extreme CE events (ECE). Where solar irradiance exceeds the expected
irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}). Analogous to
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} we have
computed the distribution of the number of occurrences of ECE events by
month in Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}.
The most active period for ECE events is in the spring and the start of
the summer (March -- June), followed by a period in the late fall
(September and October). This is probably related to the weather
characteristics of these periods, where there are continuous
alternations between clear sky periods and clouds.

@@

\caption{Seasonal variability of the number of ECE events normalized to the maximum occurring in March, in the form of a box and whisker plot. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values.}\label{fig:relative-month-occurancies-ECE}
\end{figure}

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}), shows that
there are rare cases where the OIR can exceed the TSI even more than
\(400\,\text{W}/\text{m}^2\), with the \(75\,\%\) of the cases to be
below \(200\,\text{W}/\text{m}^2\). Those finds are in accordance with
results form \citet{Vamvakas2020}, with the difference that the OIR
values are lower for Thessaloniki.

@@

\caption{Distribution of ECE above clear sky threshold, for cases that are exceeding the TSI.}\label{fig:P-extreme-distribution}
\end{figure}

@@

By creating a clear sky approximation of the GHI, which represents the
long- and short-term variation of the expected clear sky GHI, we were
able to identify cases of CE events. After analyzing the CE cases, we
found an increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\), with
the mean annual total energy of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kJ}/\text{year}\). The most active season of CE
events over Thessaloniki is concentrated on early summer, on May and
June. We found that continuous CE events can last up to \(140\) minutes,
and \(80\,\%\) of them bellow \(5\) minutes.

The magnitude of the ECE events does not exceed the values reported for
sites with more favourable conditions for the phenomenon. We have
observed cases exceeding the TSI by \(400\,\text{W}/\text{m}^{2}\) with
\(75\,\%\) of the cases under \(200\,\text{W}/\text{m}^{2}\). The
climatological characteristic of the ECE events, showed that the most
active months are spread in half of the year (March -- June, September
and October). We found that CE conditions, can have a duration of more
than an hour in rare cases, with the bulk of the cases having a duration
under 5 minutes. Some of the characteristics of CE and ECE events we
analysed have strong similarities with results by \citet{Vamvakas2020}
for the city of Patras, south of Thessaloniki, with analog differences
on the intensity of the solar radiation.

An interpretation of the CE trends shows that the interaction of GHI
with the clouds, through this 30-year period, is a dynamic phenomenon
that needs further investigation. Although, to approach it, we need more
long-term observations of AOD and clouds characteristics.

>>>>

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with lower
than average load of aerosols and humidity, which are the main factors
that attenuate the GHI, excluding clouds. With this approach the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. To compensate for this, we
defined the following threshold \(E\) to compare the measured
\(\text{GHI}\): \begin{equation}
\text{CE} : E > 15 + 1.04 \cdot \text{GHI}_\text{ref} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation}

This is the criterion of our CE identification. The constant terms were
determined through the implementation of an empirical method with manual
inspection of the CE identification results on selected days of the
whole dataset. We tested seven sets of days with different
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of EC events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of EC events, (h) randomly selected days, and (i) manually selected
days. For the latter case and where needed, we also used images from the
sky-camera to further aid the decision of the manual inspection.

The definition of the CE events with this method has a degree of
subjectivity, since the actual clear sky irradiance is not known and can
only be approximated. However, this method was proven capable in
detecting all major CE events. Where some CE events with very low OIR
may be not detected, these are few with small over-irradiance, and it is
unlikely that will affect significantly our results.

@@

\caption{Example of CE identification in Thessaloniki for 2019-07-11. The green line with blue symbols depicts the measured GHI in one minute steps. The red line shows the modelled threshold $E$ for the detection of CE events, which are denoted with red circles. The black curve represents the TSI at the top of the atmosphere, adjusted for the actual Sun-Earth distance and multiplied by the cosine of SZA.}\label{fig:example-day}
\end{figure}

@@

\caption{Example scatter plot of the measured GHI and the reference clear sky irradiance in Thessaloniki for the year 2005. The over-irradiance for CE and ECE events is color coded, while the remaining data points are shown in black.}\label{fig:example-year}
\end{figure}

@@

One aspect of this study is to investigate the time evolution of the CE
events by analyzing the GHI measurements at Thessaloniki. Cloud
enhancements can be influenced by different factors, such as the
geometry, size and optical thickness of clouds, their height in the
atmosphere and local weather regimes
\citep{Mol2023, Veerman2022, Gristey2022, Tzoumanikas2016}. Some of
these factors are related to changes in climate; hence it would be
reasonable to expect contributing to the frequency of occurrence of CE
events over Thessaloniki, as well as to the average OIR and excess
irradiation. The long-term trends were calculated using a first-order
autoregressive model with the `maximum likelihood' fitting method
\citep{Gardner1980, Jones1980}, by implementing the function `arima'
from the library `stats' of the R programming language \citep{RCT2023}.
All trends are reported together with their \(2\sigma\) error.

@@

Next we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the
monthly box and whisker plot of the CE number of occurrences normalized
with the highest median value, which occurs in June, depicting a clear
seasonal cycle. Although CE events are present throughout the year, the
most active months are May and June. During the winter (December --
February), the number of CE cases is about \(25\,\%\) of the maximum,
while in the intermediate months, the number of occurrences gradually
ramps between the maximum and minimum. This seasonality is a combined
effect of different factors, among them the types of clouds, their
frequency of occurrence, the seasonally varying relative position of the
sun, as well as the local landscape characteristics that may influence
the formation of the clouds. Unfortunately, lack of detailed data on
cloud formation, type and location is not allowing further analysis. The
interannual variability of the monthly CE events is quite high as
manifested by the large monthly extremes, especially in the summer.

@@

\caption{Seasonal variability of the number of CE events in Thessaloniki for the period 1992 -- 2023 normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have also been normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles, the thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values.}\label{fig:relative-month-occurrences}
\end{figure}

@@

\caption{Relative frequency distribution of CE events in Thessaloniki for the period 1992 -- 2023 as a function of OIR.}\label{fig:ovir-distribution}
\end{figure}

@@

The duration of the CE events is variable and can last for several
minutes or even more than an hour. In order to study the characteristics
of these consecutive events, we grouped them into bins of increasing
duration in steps of one minute. We have identified 28468 groups of CE
in the whole period of study, where the group of the longest duration of
140 minutes occurred on 07~July 2013.
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the
frequency distribution of the CE events according to their duration. We
conclude that although some groups of events last for more than an hour,
about \(80\,\%\) have duration of less than 5 minutes.

@@

\caption{Relative frequency distribution of CE groups of consequent CE events according to their duration for Thessaloniki in the period 1992 -- 2023.}\label{fig:ceg-duration-distribution}
\end{figure}

The relation between the duration and the mean OIR of the groups has
also been studied (Figure\nobreakspace{}\ref{fig:group-2d}). Evidently,
evens of high excess irradiation have small duration and vice versa. The
vast majority of grouped events are associated with small excess
irradiation (e.g., \(<5\,\text{kJ}/\text{m}^2\)) and small duration
(e.g., \(<5\,\text{min}\)) while groups with strong excess irradiation
and long duration are very rare. Similar results of this relation have
been reported by \citet{Zhang2018} in a study using a far higher
sampling rate (\(100\,\text{Hz}\)) than ours.

@@

\caption{Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1992 -- 2023. The logarithmic color scale denotes the frequency of the respective groups of events.}\label{fig:group-2d}
\end{figure}

@@

An aspect of the CE events that is commonly reported and has some
significance on the solar energy production infrastructure are the
extreme CE events (ECE), where solar irradiance exceeds the expected
irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}). Analogous to
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} we have
computed the distribution of the number of occurrences of ECE events by
month in Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}.
The most active period for ECE events is in spring and early summer
(March -- June), followed by a period in late autumn (September and
October). This is probably related to the weather characteristics in
these periods, with frequent alternations between clear sky periods and
clouds.

@@

\caption{Seasonal variability of the number of ECE events in Thessaloniki for the period 1993 -- 2023 normalized to the maximum occurring in March, in the form of a box and whisker plot. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values.}\label{fig:relative-month-occurancies-ECE}
\end{figure}

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}) shows that in
rare cases the OIR can exceed the TSI even by more than
\(400\,\text{W}/\text{m}^2\), while in \(75\,\%\) of the cases the OIR
is below \(200\,\text{W}/\text{m}^2\). The OIR for the most frequent
ECEs ranges between \(140\) and \(180\,\text{W}/\text{m}^2\). These
findings are in accordance with the results of \citet{Vamvakas2020}, the
only difference being that the OIR values reported for are higher than
those for Thessaloniki.

@@

\caption{Distribution of ECE events in Thessaloniki for the period 1992 -- 2023.}\label{fig:P-extreme-distribution}
\end{figure}

@@

By creating a clear sky approximation representing the long- and
short-term variability of the expected clear sky GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1993 -- 2023.
After analyzing the CE cases, we found an increase of
\(45.6\pm 21.9\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.

We have observed ECE cases exceeding the TSI by
\(400\,\text{W}/\text{m}^{2}\) with \(75\,\%\) of the cases under
\(200\,\text{W}/\text{m}^{2}\). The climatological characteristics of
the ECE events showed that the most active months are spread in half of
the year and particularly in the periods March -- June, September and
October. The magnitude of the ECE events identified in Thessaloniki
events does not exceed the values reported for sites with more
favourable conditions for the phenomenon \citep[e.g.][]{Cordero2023}.
Some of the characteristics of CE and ECE events we analyzed have strong
similarities with the results of \citet{Vamvakas2020} for the city of
Patras, south of Thessaloniki, albeit with differences in the magnitude
of OIR.

An interpretation of the CE trends shows that the interaction of GHI
with the clouds, through this 30-year period, is a dynamic phenomenon
that needs further investigation. For such studies it would be essential
to have more detailed information on cloud characteristics, especially
in order to investigate possible associations of the observed trends
with changes in climate.
